---
title: "Reproducible Research: Peer Assessment 1"
author: "Deniz Tekin"
date: "4/15/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setoptions, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE, fig.path='Figs/')

```

## Loading and preprocessing the data

```{r preprocess}
data <- read.csv("activity.csv") 
head(data)
summary(data)
```

## What is mean total number of steps taken per day? 

```{r}
totalstep_each_day  <- aggregate(data$steps, by=list(data$date), FUN= sum, na.rm = TRUE)
colnames(totalstep_each_day) <- c("date", "steps")
head(totalstep_each_day)
hist(totalstep_each_day$steps, main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", col = "blue", breaks = 9 )
```

### Mean and median number of steps taken each day

```{r mean_med}
mean_steps <- mean(totalstep_each_day$steps)
med_steps <- median(totalstep_each_day$steps)
mean_steps
med_steps
```

## What is the average daily activity pattern?

### Time series plot of 5-minute interval and the average number of steps taken, averaged across all days0
```{r}
average_perday <- aggregate(data$steps, by= list(data$interval), FUN = mean, na.rm = TRUE)
colnames(average_perday) <- c("intervals", "average")
head(average_perday)
plot(average_perday$intervals, average_perday$average, main = "Average Daily Activity Pattern", 
     type = "l", ylab = "Average number of steps", xlab = "Intervals", col= "red")
```

### Maximum number of steps in the 5-minute interval

```{r max}
average_perday$intervals[which.max(average_perday$average)]
```

## Imputing missing values

### Total number of missing values in the dataset

```{r}
num_na <- nrow(data[is.na(data$steps),])
num_na
```

### Filling missing values in the dataset

```{r filling}
fill <- average_perday$average[match(data$interval, average_perday$intervals)]
head(fill)
```

### Create a new dataset with the missing data filled

```{r newdata}
new_data <- transform(data, steps= ifelse(is.na(data$steps), yes = fill, no = data$steps))
head(new_data)
summary(new_data)
```

### Make a new histogram of total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.

```{r}
new_totalstep_perday  <- aggregate(new_data$steps, by=list(new_data$date), FUN= sum)
colnames(new_totalstep_perday) <- c("date", "steps")
head(new_totalstep_perday)
hist(new_totalstep_perday$steps,main = "Total number of steps taken per day", 
     xlab = "Total steps taken per day", col = "brown", breaks = 9  )
new_means <- mean(new_totalstep_perday$steps)
new_meds <- median(new_totalstep_perday$steps)
new_means
new_means
mean_steps
med_steps
```

Above we see the mean and median of total number of steps taken per day in the new dataset. They differ from the mean and median of total number of steps taken per day in the old data set because we change missing values with the mean for that 5-minute interval.

## Are there differences in activity patterns between weekdays and weekends?

```{r weekend}
new_data$date <- as.POSIXct(new_data$date, format = "%Y-%m-%d")
days <- weekdays(new_data$date)
for (i in 1:length(days))
{
    if(days[i] == "Sunday" || days[i] == "Saturday")
      {days[i] = "Weekend"}
    else
    {days[i] = "Weekday"}
}
new_data$days <- as.factor(days)
summary(new_data)
```

### Time series plot: 5-minute interval and the average number of steps taken, averaged across all  weekend days

```{r plot}
week_av <- aggregate(new_data$steps ~ new_data$interval + new_data$days,  FUN = mean)
head(week_av)

library(lattice)
xyplot(week_av$`new_data$steps` ~
week_av$`new_data$interval`|week_av$`new_data$days`, main="Average Steps by Interval",xlab="Interval", ylab="Number of Steps",layout=c(1,2), type="l", col= "red")

```











